A Deterministic and Terminating Semantics for the Synchronous Programming Language~\CEU

### 2018-02-08 01-intro.tex
Revisão até antes da semântica.

\CEU~\cite{ceu.sensys13,ceu.tecs17} is a Esterel-based~\cite{esterel.ieee91}
programming language for embedded soft real-time systems that aims to offer a
concurrent, safe, and expressive alternative to C with the characteristics that
follow:
%
\begin{description}
\item [Reactive:] code only executes in reactions to events.
\item [Structured:] programs use structured control mechanisms, such as
    \code{await} (to suspend a line of execution), and \code{par} (to combine
    multiple lines of execution).
\item [Synchronous:] reactions run atomically and to completion on each line of
    execution, i.e., there's no implicit preemption or real parallelism.
\end{description}
%
Structured reactive programming let developers write code in direct/sequential
style, recovering from the inversion of control imposed by event-driven
execution~\cite{rp.deprecating,rp.rescala,sync_async.cooperative}.
%
Synchronous languages offer a simple run-to-completion model of execution that
enable deterministic execution and make formal reasoning tractable.
For this reason, it has been successfully adopted in safety-critical real-time
embedded systems~\cite{rp.twelve}.

@@

\CEU inherits the synchronous and imperative mindset of Esterel but adopts a
simpler semantics with fine-grained execution control.~\cite{ceu.tecs17}
%
The list that follows summarizes the semantic peculiarities of \CEU:
%
\begin{itemize}
    \item Fine-grained, intra-reaction deterministic execution, which makes
          \CEU fully deterministic.
    \item Stack-based execution for internal events, which provides a limited
          but memory-bounded form of subroutines.
    \item Finalization mechanism for lines of execution, which makes abortion
          safe with regards to external resources.
\end{itemize}

>>>>

\CEU~\cite{ceu.sensys13,ceu.tecs17} is a Esterel-based~\cite{esterel.ieee91}
programming language for embedded soft real-time systems that aims to offer a
concurrent, safe, and expressive alternative to C with the characteristics that
follow:
%
\begin{description}
\item [Reactive:] code only executes in reactions to events.
\item [Structured:] programs use structured control mechanisms, such as
    \code{await} (to suspend a line of execution), and \code{par} (to combine
    multiple lines of execution).
\item [Synchronous:] reactions run atomically and to completion on each line of
    execution, i.e., there's no implicit preemption or real parallelism.
\end{description}
%
Structured reactive programming lets developers write code in direct style,
recovering from the inversion of control imposed by event-driven
execution~\cite{rp.deprecating,rp.rescala,sync_async.cooperative}.
%
Synchronous languages offer a simple run-to-completion execution model that
enables deterministic execution and make formal reasoning tractable.
For this reason, it has been successfully adopted in safety-critical real-time
embedded systems~\cite{rp.twelve}.

@@

\CEU inherits the synchronous and imperative mindset of Esterel but adopts a
simpler semantics with fine-grained execution control~\cite{ceu.tecs17}.
%
The list that follows summarizes the semantic peculiarities of \CEU:
%
\begin{itemize}
    \item Fine-grained, intra-reaction deterministic execution, which makes
          \CEU fully deterministic.
    \item Stack-based execution for internal events, which provides a limited
          but memory-bounded form of subroutines.
    \item Finalization mechanism for lines of execution, which makes abortion
          safe with regards to external resources.
\end{itemize}

### 2018-02-08 02-ceu.tex
Revisão até antes da semântica.

\CEU is a synchronous reactive language in which programs advance in a sequence
of discrete reactions to external events.
%
It is designed for control-intensive applications, supporting concurrent lines
of execution, known as \emph{trails}, and instantaneous broadcast communication
through events.
%
Computations within a reaction (such as expressions, assignments, and
system calls) are also instantaneous in accordance to the synchronous
hypothesis~\cite{rp.hypothesis}.
%
\CEU provides an \code{await} statement which is the only that actually
``consumes'' time.
%
An \code{await} statement blocks the current running trail allowing the program
to advance its other trails; when all trails are blocked, the reaction
terminates and control returns to the environment.

@@

\bgroup
\def\T<#1>{\langle\mathit{#1}\rangle}
\def\C#1#2{\hfill\rmfamily\itshape\makebox[#1\columnwidth][l]{//~#2}}
\begin{lstlisting}[
  basicstyle=\ttfamily\footnotesize,
  caption={The concrete syntax of \CEU.},
  label={lst.syntax},
]
|\C{1.}{Declarations}|
input $\T<type>$ $\T<id>$; | \C{.6}{declares external input event} |
event $\T<type>$ $\T<id>$; | \C{.6}{declares internal event}       |
var   $\T<type>$ $\T<id>$; | \C{.6}{declares variable}             |

|\C{1.}{Event handling}|
$\T<id>$ = await $\T<id>$;   | \C{.6}{awaits event and assigns received value} |
$\T<id>$ = await $\T<time>$; | \C{.6}{awaits time and assigns delayed delta}   |
emit $\T<id>$($\T<exp>$);    | \C{.6}{emits event passing value}               |

@@

|\C{1.}{Assignment \& Integration with C}|
$\T<id>$ = $\T<exp>$; | \C{.6}{assigns value to variable}                |
_$\T<id>$($\T<exps>$) | \C{.6}{calls C function (id starts with `\_'\,)} |
\end{lstlisting}
\egroup

Listing~\ref{lst.ceu} shows a complete example in \CEU that blinks a LED
whenever a radio packet is received, terminating with a button press always
with the LED off.
%
The implementation first declares the \code{BUTTON} and \code{RADIO\_RECV} as
input events (ln.~1--2).
Then, it uses a \code{par/or} composition to run two activities in parallel:
a single statement that waits for a button press before terminating (ln.~4),
and an endless loop that blinks the LED on and off (ln.~9--14).
The \code{finalize} clause (ln.~6--8) ensures that, no matter how its enclosing
trail terminates, the LED will be unconditionally turned off (ln.~7).

The \code{par/or} composition, which stands for a \emph{parallel-or}, provides
an orthogonal abortion mechanism~\cite{esterel.preemption} in which its
composed trails do not know when and how they are aborted (i.e., abortion is
external to them).
%
This is possible to do safely in synchronous languages due to the accurate
control of concurrent activities, i.e., in between every reaction, the whole
system is idle and consistent~\cite{esterel.preemption}.
%
The finalization mechanism extends orthogonal abortion to also work with
activities that use stateful resources from the environment (such as files and
network handlers), as we discuss in Section~\ref{sec.ceu.fin}.
%

\begin{lstlisting}[
  numbers=left,
  basicstyle=\ttfamily\footnotesize,
  float=h,
  caption={A program in \CEU that blinks a LED every second, terminating on a
           button press in a consistent state.},
  label={lst.ceu},
]
input void BUTTON;
input void RADIO_RECV;
par/or do
    await BUTTON;
with
    finalize with
        _led(0);
    end
    loop do
        _led(1);
        await RADIO_RECV;
        _led(0);
        await RADIO_RECV;
    end
end
\end{lstlisting}

@@

\CEU defines time as a discrete sequence of reactions to unique external
input events.
%
External input events are received from the environment, and each delimits a
new logical unit of time that triggers an associated reaction.
%
The life-cycle of a program in \CEU can be summarized as
follows~\cite{ceu.sensys13}:
%
\begin{enumerate}[i]
\item The program initiates a ``boot reaction'' in a single trail (parallel
      constructs may create new trails).
\item Active trails execute until they await or terminate, one after
      another.  This step is called a \emph{reaction chain}, and always runs in
      bounded time.
\item When all trails are blocked, the program goes idle and the environment
      takes control.
\item On the occurrence of a new external input event, the environment
      awakes \emph{all} trails awaiting that event, and the program goes back to
      step~(i).
\end{enumerate}

@@

In practice, the synchronous hypothesis for \CEU holds if reactions execute
faster than the rate of incoming input events.
%
Otherwise, the program would continuously accumulate delays between physical
occurrences and actual reactions for the input events.
%
In the soft real-time systems targeted by \CEU (e.g., sensor networks,
multimedia systems, interactive games, etc.) such delay and postponed
reactions might be tolerated by users as long as they are infrequent and the
application does not take too long to catch up with real time.
%
Note that the synchronous semantics is the norm in typical event-driven
systems, such as event dispatching in UI toolkits, game loops in game engines,
and clock ticks in embedded systems.

@@

\begin{lstlisting}[
  numbers=left,
  basicstyle=\ttfamily\footnotesize,
  float=h,
  caption={A \CEU program with a ``subroutine''.},
  label={lst.sub},
]
event int* inc;         // declares subroutine "inc"
par/or do
    var int* p;
    every p in inc do   // implements subroutine with an event iterator
        *p = *p + 1;
    end
with
    var int v = 1;
    emit inc(&v);       // calls "inc"
    emit inc(&v);       // calls "inc"
    _assert(v==3);      // asserts result after returns
end
\end{lstlisting}
\vskip-\baselineskip

@@

\CEU also supports nested \code{emit} invocations for internal events.
%
For instance, the body of the subroutine \code{inc} in Listing~\ref{lst.sub}
could \code{emit} another internal event after awaking (ln.~4), creating a
new level in the stack.
%
We can think of the stack as a record of the nested, fine-grained internal
reactions that happen inside the same outer reaction to a single external
event.

This form of subroutines has a significant limitation though: it cannot
express recursion, since an \code{emit} to itself is always ignored as a
running trail cannot be waiting on itself.
%
That being said, it is this very limitation that brings important safety
properties to subroutines.
%
First, they guaranteed to react in bounded time.
%
Second, memory for locals is also bounded, not requiring data stacks.

At first sight, event iteration such as ``\code{every e do <...> end}'' seems
to be equivalent to ``\code{loop do await e; <...> end}''.
However, the \code{loop} variation would not compile because it does not
contain a path to an external input \code{await} (\code{e} is an internal
event).
%
Event iterators have syntactic restrictions and cannot contain \code{await} or
\code{break} statements.
The absence of \code{break} guarantees that an iterator never terminates from
itself, essentially behaving as safe blocking point in the program.
%
For this reason, the restriction that execution paths within loops must
contains at least one external \code{await} is extended to contain an
\code{every} statement alternatively.

@@

\begin{figure}[h]
\begin{minipage}[h]{0.45\linewidth}
\begin{lstlisting}[numbers=right]
input void A;
input void B;
var int x = 1;
par/and do
    await A;
    x = x + 1;
with
    await B;
    x = x * 2;
end
\end{lstlisting}
\centering\small{\ax Accesses to \code{x} are never concurrent.}
\end{minipage}
%
\begin{minipage}[h]{0.53\linewidth}
%\begin{lstlisting}
\begin{lstlisting}[xleftmargin=2em]
input void A;
//
var int y = 1;
par/and do
    await A;
    y = y + 1;
with
    await A;
    y = y * 2;
end

@@

In the example \bx, the two assignments to \code{y} are simultaneous because
they execute in reaction to the same event \code{A}.
Since \CEU employs lexical order for intra-reaction statements, the execution
is still deterministic, and \code{y} always becomes \code{4} (\code{(1+1)*2}).
%
However, note that an apparently innocuous change in the order of trails
modifies the behavior of the program.
%
To mitigate this threat, \CEU performs concurrency checks at compile time to
detect conflicting accesses to shared variables:
if a variable is written in a trail segment, then a concurrent trail segment
cannot read or write to that variable~\cite{ceu.sensys13}.
%
Nonetheless, the static checks are optional and do not affect the semantics of
the language.

@@

However, aborting lines of execution that deal with external resources may lead
to inconsistencies.
%
For this reason, \CEU provides a \code{finalize} construct to unconditionally
execute a series of statements even if the enclosing block is aborted.

>>>>

\CEU is a synchronous reactive language in which programs advance in a sequence
of discrete reactions to external events.
%
It is designed for control-intensive applications, supporting concurrent lines
of execution, known as \emph{trails}, and instantaneous broadcast communication
through events.
%
Computations within a reaction (such as expressions, assignments, and
system calls) are also instantaneous considering the synchronous
hypothesis~\cite{rp.hypothesis}.
%
%\CEU provides an \code{await} statement which is the only that actually
%``consumes'' time.
%
\CEU provides an \code{await} statement that blocks the current running trail
allowing the program to advance its other trails; when all trails are blocked,
the reaction terminates and control returns to the environment.

@@

\bgroup
\def\T<#1>{\langle\mathit{#1}\rangle}
\def\C#1#2{\hfill\rmfamily\itshape\makebox[#1\columnwidth][l]{//~#2}}
\begin{lstlisting}[
  basicstyle=\ttfamily\footnotesize,
  caption={The concrete syntax of \CEU.},
  label={lst.syntax},
]
|\C{1.}{Declarations}|
input $\T<type>$ $\T<id>$; | \C{.6}{declares an external input event} |
event $\T<type>$ $\T<id>$; | \C{.6}{declares an internal event}       |
var   $\T<type>$ $\T<id>$; | \C{.6}{declares a variable}              |

|\C{1.}{Event handling}|
$\T<id>$ = await $\T<id>$;   | \C{.6}{awaits an event and assigns the received value} |
emit $\T<id>$($\T<exp>$);    | \C{.6}{emits an event passing a value} |

@@

|\C{1.}{Assignment \& Integration with C}|
$\T<id>$ = $\T<exp>$; | \C{.6}{assigns a value to a variable}              |
_$\T<id>$($\T<exps>$) | \C{.6}{calls a C function (id starts with `\_'\,)} |
\end{lstlisting}
\egroup

Listing~\ref{lst.ceu} shows a complete example in \CEU that toggles a LED
whenever a radio packet is received, terminating with a button press always
with the LED off.
%
The implementation first declares the \code{BUTTON} and \code{RADIO\_RECV} as
input events (ln.~1--2).
Then, it uses a \code{par/or} composition to run two activities in parallel:
a single statement that waits for a button press before terminating (ln.~4),
and an endless loop that toggles the LED on and off on radio receives
(ln.~9--14).
The \code{finalize} clause (ln.~6--8) ensures that, no matter how its enclosing
trail terminates, the LED will be unconditionally turned off (ln.~7).

The \code{par/or} composition, which stands for a \emph{parallel-or}, provides
an orthogonal abortion mechanism~\cite{esterel.preemption} in which its
composed trails do not know when and how they are aborted (i.e., abortion is
external to them).
%
%This is possible to do safely in synchronous languages due to the accurate
%control of concurrent activities, i.e., in between every reaction, the whole
%system is idle and consistent~\cite{esterel.preemption}.
%
The finalization mechanism extends orthogonal abortion to also work with
activities that use stateful resources from the environment (such as files and
network handlers), as we discuss in Section~\ref{sec.ceu.fin}.
%

\begin{lstlisting}[
  numbers=left,
  basicstyle=\ttfamily\footnotesize,
  float=h,
  caption={A program in \CEU that toggles a LED on every radio receive,
           terminating on a button press in a consistent state.},
  label={lst.ceu},
]
input void BUTTON;
input void RADIO_RECV;
par/or do
    await BUTTON;
with
    finalize with
        _led(0);
    end
    loop do
        _led(1);
        await RADIO_RECV;
        _led(0);
        await RADIO_RECV;
    end
end
\end{lstlisting}

@@

\CEU defines time as a discrete sequence of reactions to unique external
input events.
%
External input events are received from the environment, and each delimits a
new logical unit of time that triggers an associated reaction.
%
The life-cycle of a program in \CEU can be summarized as
follows~\cite{ceu.sensys13}:
%
\begin{enumerate}[i]
\item The program initiates a ``boot reaction'' in a single trail (parallel
      constructs may create new trails).
\item Active trails execute until they await or terminate, one after
      another.  This step is called a \emph{reaction chain}, and always runs in
      bounded time.
\item When all trails are blocked, the program goes idle and the environment
      takes control.
\item On the occurrence of a new external input event, the environment
      awakes \emph{all} trails awaiting that event, and the program goes back to
      step~(ii).
\end{enumerate}

@@

In practice, the synchronous hypothesis for \CEU holds if reactions execute
faster than the rate of incoming input events.
%
Otherwise, the program would continuously accumulate delays between physical
occurrences and actual reactions for the input events.
%
Considering the context soft real-time systems,
%targeted by \CEU (e.g., sensor networks, multimedia systems, interactive games, etc.)
such delays and postponed reactions might be tolerated as long as they are
infrequent and the application does not take too long to catch up with real
time.
%
Note that the synchronous semantics is the norm in typical event-driven
systems, such as event dispatching in UI toolkits, game loops in game engines,
and clock ticks in embedded systems.

@@

\begin{lstlisting}[
  numbers=left,
  basicstyle=\ttfamily\footnotesize,
  float=h,
  caption={A \CEU program with a ``subroutine''.},
  label={lst.sub},
]
event int* inc;         // declares subroutine "inc"
par/or do
    var int* p;
    every p in inc do   // implements "inc" through an event iterator
        *p = *p + 1;
    end
with
    var int v = 1;
    emit inc(&v);       // calls "inc"
    emit inc(&v);       // calls "inc"
    _assert(v==3);      // asserts result after the two returns
end
\end{lstlisting}
\vskip-\baselineskip

@@

\CEU also supports nested \code{emit} invocations, e.g., the body of the
subroutine \code{inc} (ln.~5) could \code{emit} an event targeting another
subroutine, creating a new level in the stack.
%
We can think of the stack as a record of the nested, fine-grained internal
reactions that happen inside the same outer reaction to a single external
event.

This form of subroutines has a significant limitation though: it cannot
express recursion, since an \code{emit} to itself is always ignored as a
running trail cannot be waiting on itself.
%
That being said, it is this very limitation that brings important safety
properties to subroutines.
%
First, they are guaranteed to react in bounded time.
%
Second, memory for locals is also bounded, not requiring data stacks.

At first sight, event iteration such as in ``\code{every e do <...> end}'' seems
to be equivalent to ``\code{loop do await e; <...> end}''.
However, the \code{loop} variation would not compile because it does not
contain a path to an external input \code{await} (\code{e} is an internal
event).
%
However, event iterators enforce syntactic restrictions and cannot contain
\code{await} or \code{break} statements.
The absence of \code{break} guarantees that an iterator never terminates from
itself, essentially behaving as a safe blocking point in the program.
%
For this reason, the restriction that execution paths within loops must
contain at least one external \code{await} is extended to alternatively contain
an \code{every} statement.

@@

\begin{figure}[h]
\begin{minipage}[h]{0.45\linewidth}
\begin{lstlisting}[numbers=right]
input void A;
input void B;
var int x = 1;
par/and do
    await A;
    x = x + 1;
with
    await B;
    x = x * 2;
end
\end{lstlisting}
\centering\small{\ax Accesses to \code{x} are never concurrent.}
\end{minipage}
%
\begin{minipage}[h]{0.53\linewidth}
%\begin{lstlisting}
\begin{lstlisting}[xleftmargin=2em]
input void A;
// (empty line)
var int y = 1;
par/and do
    await A;
    y = y + 1;
with
    await A;
    y = y * 2;
end

@@

In the example \bx, the two assignments to \code{y} are simultaneous because
they execute in reaction to the same event \code{A}.
Since \CEU employs lexical order for intra-reaction statements, the execution
is still deterministic, and \code{y} always becomes \code{4} (\code{(1+1)*2}).
%
However, note that an apparently innocuous change in the order of trails
modifies the behavior of the program.
%
To mitigate this threat, \CEU performs concurrency checks at compile time to
detect conflicting accesses to shared variables:
if a variable is written in a trail segment, then a concurrent trail segment
cannot read or write to that variable~\cite{ceu.sensys13}.
%
Nonetheless, the static checks are optional and are not a prerequisite for the
deterministic semantics of the language.

@@

However, aborting lines of execution that deal with external resources may lead
to inconsistencies.
%
For this reason, \CEU provides a \code{finalize} construct to unconditionally
execute a series of statements even if the enclosing block is externally
aborted.

### 2018-02-08 index.tex
Revisão até antes da semântica.

\title[A Deterministic and Terminating Semantics for \CEU]
{A Deterministic and Terminating Semantics
  for the Synchronous Programming Language~\CEU}

@@

\begin{abstract}
\CEU is a synchronous programming language for embedded soft real-time systems.
%
It focus on control-flow safety features, such as safe shared-memory
concurrency and safe abortion of lines of execution, while enforcing internal
determinism and time-bounded reactions to the environment.
%
In this work, we present a small-step structural operational semantics for
\CEU and a proof that reactions are deterministic and always terminate:
%
For a given timeline of input events, multiple executions of the same program
always react in bounded time and arrive at the same final memory state.
%
%We also discuss some equivalence results and the relation between the proposed
%semantics and its actual implementation.
\end{abstract}

>>>>

\title[A Memory-Bounded, Deterministic, and Terminating Semantics for \CEU]
{A Deterministic and Terminating Semantics
  for the Synchronous Programming Language~\CEU}

@@

\begin{abstract}
\CEU is a synchronous programming language for embedded soft real-time systems.
%
It focus on control-flow safety features, such as safe shared-memory
concurrency and safe abortion of lines of execution, while enforcing
memory-bounded, deterministic, and terminating reactions to the environment.
%
In this work, we present a small-step structural operational semantics for
\CEU and a proof that reactions have the properties enumerated above:
%
that for a given arbitrary timeline of input events, multiple executions of the
same program always react in bounded time and arrive at the same final finite
memory state.
%
%We also discuss some equivalence results and the relation between the proposed
%semantics and its actual implementation.
\end{abstract}
